Static Analysis by Path Focusing

### 2011-06-13 report/report.tex
report

\usepackage{verimagstudent}
\usepackage{mathcomp,amssymb}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[toc,page]{appendix} 
% DM: verimagstudent charge hyperref, et hyperref doit habituellement
% DM: passer en dernier
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}

@@

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aims at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

A fundamental fact in static analysis is that it cannot be perfectly precise:
either the set of possible states we compute is unsound (it doesn't contain all
the possible states during execution), or it is incomplete (it contains
states that can never be reached during execution). 
Abstract Interpretation is sound but incomplete, since it always computes an
upper-approximation of the set of possible states. The challenge is then to be
as precise as possible, so that we can prove properties about the program.

@@

\section{Introductory example}

@@

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}
% DM: d'habitude on met des ' sur les variables de l'etat d'arrivee
% DM: et pas de depart
We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more new elements in the
sets\dots
\end{itemize}

@@

The concrete semantics is a monotonic function $\Phi: C \rightarrow C$.
One can choose an abstract semantics $\Phi^\#: A \rightarrow A$ satisfying this
condition:

@@

\begin{proposition}
Suppose $C$ is a complete lattice, and $\Phi$ is increasing from $C$ to $C$.
Each $x \in A$, satisfying the condition $\Phi^\#(x) = x$, is an abstraction of
the least fixpoint of $\Phi$.
\end{proposition}

Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
= \gamma \circ \Phi^\#(x)$. Abstract interpretation aims at computing
such an~$x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& \Phi^\#(x_n)
\end{array}
\right.
$$
where $\perp$ is the least element of $A$.
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
\widening x_2$. This guarantees the correctness of the result.
% DM: souvent n'est defini que pour x1<=x2 ce qui oblige a utiliser
% DM: une union abstraite juste avant
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i \geq 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

\cite{Monniaux_HOSC09} gives a more general definition of the widening operator.
% DM: c'est gentil de me citer!

@@

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i \geq 0
}$$

@@

For instance, consider the program:
\begin{C}
while (x <= 50) {
	x++;
	y++;
}
\end{C}
If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
widening (which gives the result $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
a widening with the threshold $x \leq 50$ gives 
$P \widening_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
with threshold helps to find better invariants.
\end{itemize}

>>>>

\usepackage{verimagstudent}
\usepackage{mathcomp,amssymb}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[toc,page]{appendix} 
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}

@@

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aimed at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

A fundamental fact in static analysis is that it cannot be perfectly precise:
either the analysis is unsound (the set of possible states we compute does not contain all
the possible states during execution), or is incomplete (the computed set
contains states that can never be reached during execution). 
Abstract Interpretation is sound but incomplete, since it always computes an
upper-approximation of the set of possible states. The challenge is then to be
as precise as possible, so that we can prove properties about the program.

@@

\section{Introductive example}

@@

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}
We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more new elements in the
sets\dots
\end{itemize}

@@

The concrete semantics is a function $\Phi: C \rightarrow C$.
In the introductive example, this function $\Phi$ is the following:

$$\Phi\left[ \begin{array}{c}
Y_1 \\
Y_2 \\
Y_3 \\
Y_4
\end{array} \right] = \left[ \begin{array}{cc} 
x\ |& x \in \R \\
x\ |& x = 0 \vee \exists\  x'\in Y_2, x = x'+1 \\
x\ |& x \in Y_1 \wedge x < 100 \\
x\ |& x \in Y_1 \wedge x \geq 100 
\end{array} \right]$$

One can choose an abstract semantics $\Phi^\#: A \rightarrow A$ satisfying this
condition:

@@

\begin{proposition}
Suppose $C$ is a complete lattice, and $\Phi$ is increasing from $C$ to $C$.
Each $x \in A$, satisfying the condition $\Phi^\#(x) \subseteq x$, is an abstraction of
the least fixpoint of $\Phi$.
\end{proposition}

Indeed, if $\Phi^\#(x) \subseteq x$, then $\Phi \circ \gamma (x) 
\subseteq \gamma \circ \Phi^\#(x) \subseteq \gamma(x)$.
Abstract interpretation aims at computing
such an~$x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& x_n \sqcup \Phi^\#(x_n)
\end{array}
\right.
$$
where $\perp$ is the least element of $A$, and $\sqcup$ is an operator verifying
$x_1 \cup x_2 \subseteq x_1 \sqcup x_2$ and $x_1 \sqcup x_2 \in A$.
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
\widening x_2$. This guarantees the correctness of the result. Most of the time,
this operator is only defined for $x_1 \leq x_2$. In this case, we use $x_1
\widening (x_1 \sqcup x_2)$ instead.
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i \geq 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

\cite{Monniaux_HOSC09} gives a more general definition of the widening operator.

@@

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \widening\ ( x'_i \sqcup \Phi^\#(x'_i)),\ \ \forall i \geq 0
}$$

@@

For instance, consider the program:
\begin{C}
while (x <= 50) {
	x++;
	y++;
}
\end{C}
If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
we have 
\\$Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
widening (which gives the result $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
a widening with the threshold $x \leq 50$ gives 
$P \widening_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
with threshold helps to find better invariants.
\end{itemize}
